---
title: "Reproducible Research Assignment"
author: "Roshan Raghuram"
date: "July 11, 2017"
output: html_document
---

## Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

This document presents the results from Project Assignment 1 in the Coursera course Reproducible Research, written in a single R markdown document that can be processed by knitr and transformed into an HTML file.


## Load all the necessary packages

```{r}
library(knitr)
opts_chunk$set(echo =TRUE)
library(dplyr)
library(lubridate)
library(ggplot2)
```

## SETTING THE DIRECTORY AND LOADIND THE DATA
```{r}
setwd("C:/Users/ADMIN/Desktop/Data_Analytics/Reproducible Research")

data <- read.csv("activity.csv", header = TRUE, sep = ',', colClasses = c("numeric", "character",
                                                                  "integer"))
```
## TIDYING THE DATA
```{r}
data$date <- ymd(data$date)
```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

1) Calculate the total number of steps taken per day
2) If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
3) Calculate and report the mean and median of the total number of steps taken per day

#### 1)CALCULATING THE STEPS TAKEN
```{r}
steps <- data %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```

#### 2)CONSTRUCTING THE PLOT USING ggplot
```{r}
ggplot(steps, aes(x = steps)) +
  geom_histogram(fill = "firebrick", binwidth = 1000) +
  labs(title = "Histogram of Steps per day", x = "Steps per day", y = "Frequency")
```

#### 3)CALCULATING THE MEAN AND THE MEDIAN
```{r}
mean_steps <- mean(steps$steps, na.rm = TRUE)
median_steps <- median(steps$steps, na.rm = TRUE)
```

```{r}
mean_steps
```

```{r}
median_steps
```

## What is the average daily activity pattern?
1) Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
2)Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

#### CALCULATING THE AVERAGE STEPS
```{r}
interval <- data %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))
```

#### PLOTTING THE GRAPH
```{r}
ggplot(interval, aes(x=interval, y=steps)) +
  geom_line(color = "firebrick")
```

#### FINDING THE MAX STEPS:
```{r}
interval[which.max(interval$steps),]
```

## Imputing missing values
1) Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
2) Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3) Create a new dataset that is equal to the original dataset but with the missing data filled in.
4) Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

#### FILLING IN THE MISSING VALUES
```{r}
data_full <- data
nas <- is.na(data_full$steps)
avg_interval <- tapply(data_full$steps, data_full$interval, mean, na.rm=TRUE, simplify=TRUE)
data_full$steps[nas] <- avg_interval[as.character(data_full$interval[nas])]
```

#### CALCULATING THE STEPS
```{r}
steps_full <- data_full %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps)) %>%
  print
```

#### PLOTTING THE GRAPH
```{r}
ggplot(steps_full, aes(x = steps)) +
  geom_histogram(fill = "firebrick", binwidth = 1000) +
  labs(title = "Histogram of Steps per day, including missing values", x = "Steps per day", y = "Frequency")
```

#### CALCULATING THE MEAN AND THE MEDIAN
```{r}
mean_steps_full <- mean(steps_full$steps, na.rm = TRUE)
median_steps_full <- median(steps_full$steps, na.rm = TRUE)
```

```{r}
mean_steps_full
```

```{r}
median_steps_full
```

## Are there differences in activity patterns between weekdays and weekends?

* For this part the weekdays() will come handy. Use the dataset with the filled-in missing values for this part.

1) Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
2) Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

#### WHETHER WEEKDAY OR WEEKEND: 
```{r}
data_full <- mutate(data_full, weektype = ifelse(weekdays(data_full$date) == "Saturday" | weekdays(data_full$date) == "Sunday", "weekend", "weekday"))
data_full$weektype <- as.factor(data_full$weektype)
head(data_full)
```

#### AVERAGE STEPS IN 5 MINS INTERVALS:
```{r}
interval_full <- data_full %>%
  group_by(interval, weektype) %>%
  summarise(steps = mean(steps))
```

#### PLOTTING THE GRAPH:
```{r}
s <- ggplot(interval_full, aes(x=interval, y=steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow=2)
print(s)
```

